---
title: "Final"
output: html_document
date: "2024-02-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### This project explors the "county_demographics.csv" dataset. It works with demographics from counties in the United States . 

## Data preprocessing 

First, I load the necessary libraries (install packages, if needed), read the data set and print its content!
```{r }
# Install and load packages if not already installed
#install.packages("corrplot")
# install.packages("GGally")
#install.packages("cowplot")
#install.packages("ggrepel")
#install.packages("usmap")

library(usmap)
library(ggrepel)
library(cowplot)
library(GGally)
library(tidyverse)
library(dplyr)
library(ggplot2)


# Set the path to your CSV file
csv_file_path <- "county_demographics.csv"

# Read the CSV file
data <- read.csv(csv_file_path)

# Print the contents of the data
#head(data)

```

## 1. Descriptive statistics, Missing values, Outliers! 

### Descriptive statistics

Summary of dataset
```{r }
summary(data)
```


Histogram plots in order to visualize the distribution of numeric variables in the dataset.

```{r }
numeric_vars <- sapply(data, is.numeric)
numeric_data <- data[, numeric_vars]

par(mfrow=c(2,2)) # Change the layout to arrange plots in a 2x2 grid
for (col in names(numeric_data)) {
  hist(data[[col]], main=col, xlab="Value", ylab="Frequency")
}
``` 


Scatter plots in order to visualize the relationship between pairs of numeric variables!

```{r }
# Adjust the size of the plotting region
par(mfrow=c(2, 2), mar=c(4, 4, 2, 2))

# Create scatter plots
for (col in names(numeric_data)) {
  plot(data[[col]], main=col, xlab="Index", ylab="Value")
}

``` 

### Missing values handling
Missing values are denoted by "-1" on the dataset. Thus, I replace all "-1" values with "NA" so that I can handle NAs after!

I am imputing missing values using the mean value of the row (, from which the value is missing)

```{r }
# Replace -1 with NA
data[data == -1] <- NA

# Impute missing values with mean
data  <- data %>%
  mutate_all(~ifelse(is.na(.), mean(., na.rm = TRUE), .))

```



### Adding new Education Column
I am adding the column "Education.Elementary_School_and_below" and I calculate its values!
(This is done after the NA handling so that the missing values are not included in the calculation as "-1" )

```{r }

# Rename column with special characters
names(data)[names(data) == "Education.Bachelor.s.Degree.or.Higher"] <- "Education.Bachelors_Degree_or_Higher"
names(data)[names(data) == "Education.High.School.or.Higher"] <- "Education.High_School_or_Higher"


# Perform calculation
data <- data %>%
  mutate(Education_Elementary_School_and_below = 100 - (Education.Bachelors_Degree_or_Higher + Education.High_School_or_Higher))

```

### Boxplot for outlier values 
Plotting of boxplots in order to detect extreme values! 
```{r }
# Boxplot for each numeric variable
numeric_vars <- sapply(data, is.numeric)
numeric_data <- data[, numeric_vars]

par(mfrow=c(2,2)) # Change the layout to arrange plots in a 2x2 grid
for (col in names(numeric_data)) {
  boxplot(data[[col]], main=col)
}
```


## 2. Data analysis and plotting! Univariate - Multivariate analysis 
### For this task, I chose age, housing and income as the 3 categories that I will further explore! 

### First, I explore the category "age" - Univariate analysis for each of the three columns of this category


Summary of statistics for the three age-related variables!
```{r }
summary_age <- data %>%
  select(`Age.Percent.65.and.Older`, `Age.Percent.Under.18.Years`, `Age.Percent.Under.5.Years`) %>%
  summary()
print(summary_age)
```


Visualizations for each age-related variable: Histograms!

Histogram plots for each age-related variable distribution!

```{r }

#  Histogram for 'Age.Percent.65.and.Older'
ggplot(data, aes(x = `Age.Percent.65.and.Older`)) +
  geom_histogram(binwidth = 2, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Age.Percent.65.and.Older",
       x = "Percentage", y = "Frequency") +
  theme_minimal()

#  Histogram for 'Age.Percent.Under.18.Years'
ggplot(data, aes(x = `Age.Percent.Under.18.Years`)) +
  geom_histogram(binwidth = 2, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Age.Percent.Under.18.Years",
       x = "Percentage", y = "Frequency") +
  theme_minimal()

#  Histogram for 'Age.Percent.Under.5.Years'
ggplot(data, aes(x = `Age.Percent.Under.5.Years`)) +
  geom_histogram(binwidth = 2, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Age.Percent.Under.5.Years",
       x = "Percentage", y = "Frequency") +
  theme_minimal()


```


Histogram plots for each age-related variable, faceted by State
```{r }

# Histogram for 'Age.Percent.65.and.Older' faceted by State
ggplot(data, aes(x = `Age.Percent.65.and.Older`)) +
  geom_histogram(binwidth = 2, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Age.Percent.65.and.Older by State",
       x = "Percentage", y = "Frequency") +
  facet_wrap(~State) +  # Facet by State
  theme_minimal()

# Histogram for 'Age.Percent.Under.18.Years' faceted by State
ggplot(data, aes(x = `Age.Percent.Under.18.Years`)) +
  geom_histogram(binwidth = 2, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Age.Percent.Under.18.Years by State",
       x = "Percentage", y = "Frequency") +
  facet_wrap(~State) +  # Facet by State
  theme_minimal()

# Histogram for 'Age.Percent.Under.5.Years' faceted by State
ggplot(data, aes(x = `Age.Percent.Under.5.Years`)) +
  geom_histogram(binwidth = 2, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Age.Percent.Under.5.Years by State",
       x = "Percentage", y = "Frequency") +
  facet_wrap(~State) +  # Facet by State
  theme_minimal()
``` 
### Maps for the age-related features! 
```{r }
# Rename "State" to "state" so that it is recognisable by the package
data <- data %>% 
  rename(state = State)


# Create maps for each age-related feature
# Age.Percent.65.and.Older
map_65_and_older <- plot_usmap(data = data, values = "Age.Percent.65.and.Older", 
                               color = "blue") + 
                    scale_fill_continuous(name = "Percentage", label = scales::percent) +
                    labs(title = "Percentage of Population Aged 65 and Older by State")

# Age.Percent.Under.18.Years
map_under_18 <- plot_usmap(data = data, values = "Age.Percent.Under.18.Years", 
                           color = "blue") + 
                scale_fill_continuous(name = "Percentage", label = scales::percent) +
                labs(title = "Percentage of Population Under 18 Years by State")

# Age.Percent.Under.5.Years
map_under_5 <- plot_usmap(data = data, values = "Age.Percent.Under.5.Years", 
                          color = "blue") + 
               scale_fill_continuous(name = "Percentage", label = scales::percent) +
               labs(title = "Percentage of Population Under 5 Years by State")

# Display the maps
print(map_65_and_older)
print(map_under_18)
print(map_under_5)

``` 

### Then, I explore the category "housing" - Univariate and Multivariate analysis for each of the five columns of this category

Summary of statistics for the three housing -related variables!
```{r }
summary_housing <- data %>%
  select(`Housing.Homeownership.Rate`, `Housing.Households`, `Housing.Housing.Units`, `Housing.Median.Value.of.Owner.Occupied.Units`,`Housing.Persons.per.Household` ) %>%
  summary()
print(summary_housing)
```

Univariate analysis for housing-related variables: Boxplots per State! 
```{r }
# Define the housing variables
housing_variables <- c("Housing.Homeownership.Rate", "Housing.Households", 
                        "Housing.Housing.Units", "Housing.Median.Value.of.Owner.Occupied.Units",
                        "Housing.Persons.per.Household")


univariate_results <- lapply(housing_variables, function(variable) {
  # Create a boxplot for each housing variable
  boxplot(data[[variable]] ~ data$state, 
          main = paste("Boxplot of", variable, "by State"),
          xlab = "State", ylab = variable)
})


``` 

Multivariate analysis: Correlation Matrix among all housing-related variables! 
```{r }

# Correlation matrix
correlation_matrix <- cor(data[housing_variables])
library(corrplot)

# Plot correlation matrix as a heatmap
corrplot(correlation_matrix, method = "color")
```

### Maps for the housing-related features! 
```{r }
# Select the relevant housing features
housing_data <- data %>%
  select(state, Housing.Homeownership.Rate, Housing.Households,
         Housing.Housing.Units, Housing.Median.Value.of.Owner.Occupied.Units,
         Housing.Persons.per.Household)

# Define a function to create a map with specified data
create_housing_map <- function(data, column_name) {
  # Create a map with data
  plot_usmap(data = data, values = column_name) +
    scale_fill_continuous(name = column_name)
}

# Create maps for each housing feature
for (column_name in colnames(housing_data)[-1]) {
  housing_map <- create_housing_map(housing_data, column_name)
  print(housing_map)
}

``` 


### Lastly, I explore the category "income" - Univariate analysis for each of the two columns of this category

Summary of statistics for the two income-related variables!
```{r }
summary_income <- data %>%
  select(`Income.Median.Houseold.Income`, `Income.Per.Capita.Income`) %>%
  summary()
print(summary_income)
```


Univariate analysis: Histograms for the two variables

```{r }
# Univariate Analysis: Histograms for each income-related variable
ggplot(data, aes(x = Income.Median.Houseold.Income)) +
  geom_histogram(binwidth = 2000, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Median Household Income",
       x = "Income", y = "Frequency") +
  theme_minimal()

ggplot(data, aes(x = Income.Per.Capita.Income)) +
  geom_histogram(binwidth = 1000, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Per Capita Income",
       x = "Income", y = "Frequency") +
  theme_minimal()


``` 

Histograms for the two variables in relation to the "state"
```{r }
# Univariate Analysis: Histograms for each income-related variable by State
ggplot(data, aes(x = Income.Median.Houseold.Income, fill = state)) +
  geom_histogram(binwidth = 2000, color = "black") +
  labs(title = "Distribution of Median Household Income by State",
       x = "Income", y = "Frequency") +
  theme_minimal()

ggplot(data, aes(x = Income.Per.Capita.Income, fill = state)) +
  geom_histogram(binwidth = 1000, color = "black") +
  labs(title = "Distribution of Per Capita Income by State",
       x = "Income", y = "Frequency") +
  theme_minimal()
``` 

 
## 3. PCA: Visualization of states in 2d space! 
```{r pca }
# Perform PCA
pca <- prcomp(data[housing_variables], scale. = TRUE)

# Extract PCA scores
pca_scores <- as.data.frame(pca$x[, 1:2])  # Extracting the first two principal components
pca_scores$state <- data$state  # Adding the State column to the PCA scores dataframe

# Plot the PCA scores with state labels and colors
p <- ggplot(pca_scores, aes(x = PC1, y = PC2, color = state)) +
  geom_point() +
  geom_text_repel(aes(label = state), size = 3) + # Adding state labels with repelling
  labs(x = "Principal Component 1", y = "Principal Component 2", title = "PCA Visualization of States") +
  theme_minimal() +
  theme(legend.position = "none")  # Hide the legend for state colors

# Create a data frame for the legend
legend_data <- data.frame(state = unique(pca_scores$state), 
                           Color = scales::hue_pal()(length(unique(pca_scores$state))))

# Plot the legend manually
legend_plot <- ggplot(legend_data, aes(x = 1, y = seq_along(state), color = state, label = state)) +
  geom_point(size = 3) +
  geom_text(aes(label = state), hjust = -0.1) +
  scale_color_manual(values = unique(legend_data$Color)) +
  theme_void() +
  labs(color = "state")

# Arrange the main plot and legend plot side by side
combined_plot <- cowplot::plot_grid(p, legend_plot, ncol = 2, align = "h", rel_widths = c(0.8, 0.2))

# Display the combined plot
print(combined_plot)
``` 
